---
title: "Advanced Numerical 2.0"
---

```{r}
options(repos = c(CRAN = "https://cran.rstudio.com"))
```

## Prepare Environment

```{r}
install.packages("broom.mixed")
install.packages("kableExtra")
install.packages("lmerTest")
library(haven)
library(tidyverse)
library(broom.mixed)
library(here)
library(gtsummary)
library(DT)
library(kableExtra)
library(lme4)
library(lmerTest)
```

```{r}
library(readr)
Life_Expectancy_Data <- read_csv("C:/Users/H P/OneDrive - The Goose and Duck/Desktop/Data Survival/Life Expectancy Data - Latest.csv")
glimpse(Life_Expectancy_Data)
str(Life_Expectancy_Data)
```

```{r}
Life_Expectancy_Data <- Life_Expectancy_Data %>%
  mutate(across(
    .cols = where(is.character),
    .fns = as.factor
  ))
glimpse(Life_Expectancy_Data)

```

```{r}
Life_Expectancy_Data <- Life_Expectancy_Data %>%
  mutate(`Life expectancy` = as.numeric(`Life expectancy`))

Life_Expectancy_Data <- Life_Expectancy_Data %>%
  mutate(Year_since2000 = as.numeric(as.character(Year)) - 2000)

glimpse(Life_Expectancy_Data)
```

```{r}
Life_Expectancy_Data <- Life_Expectancy_Data %>%
  mutate(`Life expectancy` = as.numeric(`Life expectancy`))

Life_Expectancy_Data <- Life_Expectancy_Data %>%
  mutate(Year = as.factor(Year))

glimpse(Life_Expectancy_Data)
```

```{r}
tbl_summary(Life_Expectancy_Data)
```

```{r}
library(gtsummary)
library(dplyr)

# Create descriptive table
table_summary <- Life_Expectancy_Data %>%
  select(Status, Year, Year_since2000, `Life expectancy`) %>%
  tbl_summary(
    by = Status,
    *,
    digits = all_continuous() ~ 1
  ) %>%
  add_overall() %>%
  modify_header(label ~ "**Variable**") %>%
  modify_spanning_header(all_stat_cols() ~ "**Summary Statistics by Development Status**") %>%
  modify_caption("**Table 1. Descriptive Statistics of Life Expectancy Dataset**")

# Print the table
table_summary

```

Test

```{r}
library(ggplot2)
library(dplyr)
Life_Expectancy_Data %>%
  group_by(Status) %>%
  summarise(Avg_LE = mean(`Life expectancy`, na.rm = TRUE)) %>%
  ggplot(aes(x = Status, y = Avg_LE, fill = Status)) +
  geom_bar(stat = "identity") +
  labs(title = "Life Expectancy by Development Status",
       y = "Average Life Expectancy")

```

```{r}
ggplot(Life_Expectancy_Data, aes(x = Year, y = `Life expectancy`, color = Status)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Trends in Life Expectancy by Status")

```

```{r}
library(ggplot2)

ggplot(Life_Expectancy_Data, aes(x = Year_since2000, y = `Life expectancy`, group = Country, color = Status)) +
  geom_line(alpha = 0.4) +
  labs(title = "Life Expectancy Over Time by Country and Status",
       x = "Year", y = "Life Expectancy") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
ggplot(Life_Expectancy_Data, aes(x = Year_since2000, y = `Life expectancy`, group = Country, color = Status)) +
  geom_line(alpha = 0.2) +
  stat_summary(aes(group = Status), fun = mean, geom = "line", size = 1.2, linetype = "solid") +
  labs(title = "Life Expectancy Trends with Group Mean Averages According to Status (Developed and Developing)",
       x = "Year", y = "Life Expectancy") +
  theme_minimal()

```

```{r}
m.lm <- lm(Life_Expectancy_Data$`Life expectancy` ~ 1, data = Life_Expectancy_Data)
summary(m.lm)
```

```{r}
m0 <- 
  lmer(Life_Expectancy_Data$`Life expectancy` ~ 1 + (1 | Status/Country), 
       data = Life_Expectancy_Data, REML = FALSE)
summary(m0)
```

```{r}
tidy(m0) %>%
kbl() %>%
kable_styling()
```

```{r}
mlr <- lm(Life_Expectancy_Data$`Life expectancy` ~ 1, data = Life_Expectancy_Data)
summary(mlr)
```

```{r}
logLik(mlr) ; logLik(m0)
```

## Random intercept

```{r}
ri <- lmer(Life_Expectancy_Data$`Life expectancy` ~ Status + Year_since2000 + (1 | Country),
data = Life_Expectancy_Data,
REML = FALSE)
summary(ri)
```

```{r}
ri <- lmer(Life_Expectancy_Data$`Life expectancy` ~ Year_since2000 + (1 | Status/Country),
data = Life_Expectancy_Data,
REML = FALSE)
summary(ri)
```

```{r}
tidy(ri, conf.int = TRUE) %>%
kbl %>%
kable_styling()
```

```{r}
pred_score <- fitted(ri)
head(pred_score, 106)
```

```{r}
rand_ef <- ranef(ri)
head(rand_ef$Country, 106)
```

```{r}
ri_fitted <- augment(ri)
```

```{r}
ri_fitted %>%
slice(1590:1605)
```

```{r}
ggplot(ri_fitted, aes(x = Year_since2000, y = .fitted, group = Country, color = Status)) +
  geom_point(alpha = 0.3) +
  geom_line(alpha = 0.3) +
  ylab('Fitted Life Expectancy') +
  xlab('Year (where 0 = Year 2000)') +
  ggtitle('Fitted Values from Random Intercept Model with Cohort Year') +
  theme_bw()


```

```{r}
ggplot(ri_fitted, aes(x = Year_since2000, y = .fitted, group = Country, color = Status)) +
  geom_point(alpha = 0.3) +
  geom_line(alpha = 0.3) +
  ylab('Fitted Life Expectancy') +
  xlab('Year (where 0 = 1990)') +
  ggtitle('Fitted Values from Random Intercept Model with Cohort Year') +
  theme_bw()

```

## Random Slope

```{r}
Life_Expectancy_Data <- Life_Expectancy_Data %>%
  mutate(Year_since2000 = as.numeric(as.character(Year)) - 2000)

glimpse(Life_Expectancy_Data)
```

```{r}
rs <- lmer(`Life expectancy` ~ Status + Year_since2000 + (1 + Year_since2000 | Country),
           data = Life_Expectancy_Data,
control = lmerControl(optimizer = 'bobyqa'),
REML = FALSE)
summary(rs)
```

### Interpretation

The model estimates that the average life expectancy in the year 2000 for developed countries was **75.55 years**. Developing countries had a significantly lower baseline, by about **10.86 years**. Over time, life expectancy increases by approximately **0.35 years per year** on average across all countries.

There is substantial variation in baseline life expectancy between countries, with a **between-country variance of 86.0**, indicating that some countries start much higher or lower than others. The yearly improvement in life expectancy also varies between countries, with a **slope variance of 0.076**, showing that not all countries progress at the same rate.

Importantly, there is a **strong negative correlation (-0.71)** between a country's starting point and its growth rate. This means countries that began with higher life expectancy tend to improve more slowly, while those starting lower tend to catch up faster.

This model accounts for both country-level differences and global time trends, providing a flexible and realistic understanding of how life expectancy evolves across countries

```{r}
tidy(rs) %>% kbl() %>%
kable_styling()
```

```{r}
anova(ri, rs)
```

```{r}
rs_res <- augment(rs)
head(rs_res, 20)
summary(rs_res)
```

```{r}
ggplot(rs_res, aes(x = Year_since2000, y = .fitted, group = Country, color = Status)) +
  geom_point(alpha = 0.3) +
  geom_line(alpha = 0.3) +
  ylab('Fitted Life Expectancy') +
  xlab('Year (where 0 = Year 2000)') +
  ggtitle('Fitted Values from Random Intercept Model with Cohort Year') +
  theme_bw()
```

```{r}
ra.eff.rs <- ranef(rs, condVar = TRUE)
datatable(ra.eff.rs$Country)
plot(ra.eff.rs)
```

```{r}
ra.eff.rs.sc <- ra.eff.rs$Country
names(ra.eff.rs.sc)
```

```{r}
ra.eff.rs.sc <- ra.eff.rs.sc %>%
rename(rs_slope = Year_since2000, rs_int = "(Intercept)")
ra.eff.rs.sc %>%
ggplot(aes( x = rs_int, y = rs_slope)) +
geom_point() +
geom_vline(xintercept = 0) +
geom_hline(yintercept = 0)
```

```{r}
datatable(rs_res)
```

```{r}
ggplot(data = rs_res, aes(x = Year_since2000, y = .fitted, group = Country, color = Status)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(group = Country), alpha = 0.5) +
  labs(
    y = "Fitted Life Expectancy",
    x = "Years since 2000",
    title = "Fitted Life Expectancy by Country Over Time",
    subtitle = "From a Random Slope Model"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11)
  )

```

```{r}
install.packages("merTools")
library(merTools)
random_effects_sim <- REsim(rs)
plotREsim(random_effects_sim)
```

```{r}
library(lattice)
randoms <- ranef(rs, condVar = TRUE)
dotplot(randoms)
```

```{r}
plot(rs)
```

```{r}
library(lattice)
qqmath(rs)
```

## Test

```{r}
install.packages("rnaturalearthdata")

library(ggplot2)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# Load required libraries
library(ggplot2)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# Load world shapefile data
world <- ne_countries(scale = "medium", returnclass = "sf")

# Filter your data for most recent year (2015 = Year_since2000 15)
map_data <- Life_Expectancy_Data %>%
  filter(Year_since2000 == 15) %>%
  select(Country, `Life expectancy`) %>%
  group_by(Country) %>%
  summarise(LifeExp = mean(`Life expectancy`, na.rm = TRUE))

# Join with world map using country name
world_map <- left_join(world, map_data, by = c("name" = "Country"))

# Plot world heat map
ggplot(data = world_map) +
  geom_sf(aes(fill = LifeExp), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "lightgrey", name = "Life Expectancy (2015)") +
  theme_minimal() +
  labs(
    title = "Global Life Expectancy by Country",
    subtitle = "Year: 2015",
    caption = "Source: WHO Dataset"
  )

```

```{r}
install.packages(c("ggplot2", "dplyr", "sf", "rnaturalearth", "rnaturalearthdata", "gganimate", "transformr"))

```

```{r}
library(ggplot2)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(gganimate)

# Load world map
world <- ne_countries(scale = "medium", returnclass = "sf")

# Prepare data: average life expectancy by country and year
map_data <- Life_Expectancy_Data %>%
  group_by(Country, Year_since2000) %>%
  summarise(LifeExp = mean(`Life expectancy`, na.rm = TRUE)) %>%
  ungroup()

# Merge with world map
world_animated <- left_join(world, map_data, by = c("name" = "Country"))

# Create animated heatmap
p <- ggplot(world_animated) +
  geom_sf(aes(fill = LifeExp), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "darkgrey", name = "Life Exp") +
  theme_minimal() +
  labs(
    title = "Global Life Expectancy by Country",
    subtitle = "Year: {frame_time + 2000}",
    caption = "Source: WHO Dataset"
  ) +
  transition_time(Year_since2000) +
  ease_aes('linear')  # Smooth transition

# Animate
animate(p, nframes = 200, fps = 10, width = 900, height = 500, renderer = gifski_renderer())

# To save the animation
anim_save("life_expectancy_world_map.gif")

```

```{r}
library(dplyr)
library(ggplot2)
library(gganimate)

# Filter only selected ASEAN countries
asean_data <- Life_Expectancy_Data %>%
  filter(Country %in% c("Malaysia", "Indonesia", "Singapore", "Thailand")) %>%
  select(Country, Year_since2000, `Life expectancy`) %>%
  arrange(Country, Year_since2000)

```

```{r}
p <- ggplot(asean_data, aes(x = Year_since2000, y = `Life expectancy`, color = Country, group = Country)) +
  geom_line(size = 1.2) +
  geom_point(size = 5) +
  labs(
    title = "Life Expectancy in ASEAN Countries: Year {frame_time + 2000}",
    x = "Years Since 2000",
    y = "Life Expectancy",
    color = "Country"
  ) +
  theme_minimal(base_size = 14) +
  transition_time(Year_since2000) +
  ease_aes("linear")

# Render animation
animate(p, nframes = 100, fps = 10, width = 800, height = 500, renderer = gifski_renderer())

# Save animation
anim_save("asean_life_expectancy.gif")

```
